---
title: "Bike Sharing"
author: "Mizbah"
date: "2023-03-13"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ride Sharing Data Analysis

**The data is from an open source data platform. It has the bike sharing data of Divvy company for 2nd to 4th quarters of 2019 and the first quarter of 2020.**

```{r bikedata, warning=FALSE, message=FALSE}
options(warn = -1)
library(tidyverse)
library(lubridate)
library(ggplot2)


setwd("E:/Coursera/Google Data Analytics Professional Certificate/C8 - Google Data Analytics Capstone/Data2")
getwd()
q2_2019 <- read_csv("2019_Q2.csv")
q3_2019 <- read_csv("2019_Q3.csv")
q4_2019 <- read_csv("2019_Q4.csv")
q1_2020 <- read_csv("2020_Q1.csv")
```

### Data wrangling and combining data
```{r wrangling and join,  warning=FALSE, message=FALSE}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)

#Let's get the data to uniform format with same column name
q2_2019 <- rename(q2_2019,
                   ride_id="01 - Rental Details Rental ID",
                   rideable_type="01 - Rental Details Bike ID",
                   started_at="01 - Rental Details Local Start Time",
                   ended_at="01 - Rental Details Local End Time",
                   start_station_name = "03 - Rental Start Station Name",
                   start_station_id = "03 - Rental Start Station ID",
                   end_station_name = "02 - Rental End Station Name",
                   end_station_id = "02 - Rental End Station ID",
                   member_casual = "User Type")

q3_2019 <- rename(q3_2019
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)
                  
q4_2019 <- rename(q4_2019
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)

#Inspecting data frames and looking for inconsistencies
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)

#converting ride_id and rideable_type to character
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id),
                  rideable_type=as.character(ride_id))

q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

#Joining data
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)

#Removing lat, long, birthyear, and gender fields as they were dropped in 2020
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

#Let's go over the joined data for a bird's eye view
colnames(all_trips)
nrow(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

#### There are a few problems we will need to fix:
- In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
- The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
- We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
- There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

```{r formatting,  warning=FALSE, message=FALSE}
# In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"

all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

# Add columns that list the date, month, day, and year of each ride
# This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

# Add a "ride_length" calculation to all_trips (in seconds)
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)

#checking column structure
str(all_trips)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative. We need to get rid of this records.

``` {r remove data,  warning=FALSE, message=FALSE}
all_trips_v2<-all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
head(all_trips_v2)
```

### Let's run some descriptive anlaysis.
```{r descriptive anlaysis,  warning=FALSE, message=FALSE}
#Ride length
summary(all_trips_v2$ride_length)

#Comparing members and casual riders
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN=mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN=median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

# Average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+all_trips_v2$day_of_week, FUN = mean)

# We can see that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

# Analyzing ridership data by type and weekday.
all_trips_v2 %>% 
  mutate(weekday=wday(started_at, label=TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, weekday)

# Now, let's run the average ride time by each month for members vs casual users.
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$month, FUN = mean)

# Analyzing ridership data by type and month
all_trips_v2 %>% 
  mutate(month=month(started_at, label=TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, month)
```

### Let's do some visualization on our analysis.
```{r plotting,  warning=FALSE, message=FALSE}
# Let's visualize the average ride duration by rider type for each day.
all_trips_v2 %>% 
  mutate(weekday=wday(started_at, label=TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday, y=average_duration, fill=member_casual))+
  geom_col(position="dodge")

#Let's plot it in a line chart.
all_trips_v2 %>% 
  mutate(weekday=wday(started_at, label=TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday, y=average_duration, color=member_casual, group=member_casual))+
  geom_line()

# We see a huge difference between the ride duration between casual and subscribed members for each day.

# Let's visualize the average ride duration by rider type for each month.
all_trips_v2 %>% 
  mutate(month=month(started_at, label=TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, month) %>% 
  ggplot(aes(x=month, y=average_duration, fill=member_casual))+
  geom_col(position="dodge")

#Let's plot it in a line chart as well
all_trips_v2 %>% 
  mutate(month=month(started_at, label=TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides=n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual, month) %>% 
  ggplot(aes(x=month, y=average_duration, color=member_casual, group=member_casual))+
  geom_line()

# A huge difference also exists in monthly ride duration between member types.
```